---
title: "Inferene_2"
output: word_document
date: "2023-09-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Random Variables:

function - the output follows some distribution, or structured pattern

## properties of random variables

Concept 2.5 (Distribution):A distribution describes the ‘pattern’ that the random variable follows. Imagine a ‘recipe,’ (the distribution) that gives the instructions to make a specific ‘meal’ (the random variable).


Concept 2.6 (Expectation): 

A random variable’s expectation is another name for its average, and it is denoted by $E[X] = \sum_i x_i.P(X=x_i)$

Concept 2.7 (Variance): As the name overtly suggests, this describes how much spread is inherent in a certain random variable, and is notated $Var[X] = \sum_i (x_i-E[X])^2.P(X=x_i)

Concept 2.8 PMF and CDF:

the PMF gives the probability that the random variable takes on a certain value.

A step further than the PMF is the cumulative density function (which is the same for continuous and discrete variables; both have CDFs, there is no CMF!). P(X =< x)


Concept 2.9 (Support): 
The support of a distribution is simply the set of possible values that the random variable can take on. For example, if you had a random variable that could only spit out negative numbers or 0, then the support would be negative infinity to 0

```{r}
#replicate
set.seed(110)
sims = 1000

#generate winnings
winnings = sample(c(10, 0, -5), sims, replace = TRUE, prob = c(2/6, 3/6, 1/6))

#the mean and variance should match above (2.5 and 31.25)
mean(winnings); var(winnings)
```


## Binomial:
- discrete: 1,2,3..
- The story of the Binomial is we perform  
n
  independent trials, each with only two outcomes (usually we think of these two outcomes as success or failure) and with a probability of succes p that stays constant from trial to trial.
  
$Bin~(n,p)$
$T~Bin(30, 5)= 30.5.(1-5)= 7.7$
$E(X)=n.p$
$Var(X)= m.p.q$ q=1-p

$X~Bin(5,5) = P(X=x) = \binom{n}{x}.p^x.q^{n-x}= \binom{5}{x}.5^x.5^{5-x}$

- support - 0, n

```{r}
#find P(X = 3), where X ~ Bin(10, 1/2)
dbinom(3, 10, 1/2)

#find P(X <= 6), where X ~ Bin(15, 1/3)
pbinom(6, 15, 1/3)

#generate 5 random draws from X, where X ~ Bin(30, 1/4)
rbinom(5, 30, 1/4)
```


# Disreate Random Varibles

1. Bernoulli

2. Geometric

3. First Success

4. Negative Binomial

5. Poisson

6. Hypergometric

7. Expectation, Indicators and Memorylessness


  